---
title: "624_HW1"
date: "02/02/22"
author: "Sean Connin"
output:
  html_document: 
    toc: TRUE
    toc-title: ""
    toc_depth: 3
    toc_float: 
      collapsed: TRUE
      smooth_scroll: FALSE
    number_sections: true
    df_print: paged
    code_folding: hide
    theme: yeti
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message = FALSE, error=FALSE)
```

```{r}
library(fpp3)
library(magrittr)
library(tidyverse)
library(patchwork)
library(readxl)
library(flextable)
library(USgas)

```

# Time Series Graphics: Q1

**Use the help function to explore what the series gafa_stock, PBS, vic_elec and pelt represent.**

-   gafa_stock: Historical stock prices from 2014-2018 for Google, Amazon, Facebook and Apple. All prices are in \$USD.

-   PBS: Monthly Medicare Australia prescription data

-   elec_vic: half-hourly electricity demand for Victoria, Australia

-   pelt: Hudson Bay Company trading records for Snowshoe Hare and Canadian Lynx furs from 1845 to 1935.

**Use autoplot() to plot some of the series in these data sets.**

**What is the time interval of each series?**

-   gafa_stock: *day*
-   pbs: *month*
-   vic_elec: *1/2 hour*
-   pelt: *annual*

```{r}

# explore series

help(gafa_stock)

help(PBS)

help(vic_elec)

help(pelt)

# 1.a. Use autoplot() to plot some of the series in these data sets

#plot interannual closing price

amazon<-gafa_stock%>%    #closing daily stock price:Amazon
  filter(Symbol=='AMZN')

p1<-autoplot(amazon, Adj_Close, color='darkblue') +
  labs(title = "Amazon Adjusted Stock Closing Price",
       subtitle = "2014-2018",
       y = "Adjusted Closing Price ($)")+
  theme_classic()

# plot avg daily demand

daily_avg <- aggregate(Demand~Date,vic_elec, mean)%>% # avg daily demand: vic_elec
  as_tsibble()

p2<-autoplot(daily_avg, Demand, color='darkblue')+
  labs(title = "Average Daily Electricity Demand in Victoria, Australia",
       subtitle = " Year = 2012",
       y = "Demand (MW))")+
  theme_classic()

(p1/p2)
```

# Time Series Graphics: Q2

**Use filter() to find what days corresponded to the peak closing price for each of the four stocks in gafa_stock.**

*AMZN: \$2,039.51*

*GOOG: \$1,268.33*

*AAPL: \$232.07*

*FB: \$217.50*

```{r}

# create function and map with purrr to find peak closing dates
# note we need to convert stibble to tibble

company<-set_names(c(gafa_stock$Symbol%>%unique()))

f = function(x) {
  gafa_stock%>%
  as_tibble()%>% 
  filter(Symbol== x)%>%
  summarise(maxi=max(Close))%>%
  reduce(full_join, by = 'Symbol')
}

(max_close<-map_df(company, ~f(.x))%>%pivot_longer(cols = 1:4,names_to = "Company", values_to = "Peak_Close")%>%arrange(desc(Peak_Close))%>%flextable())


# The following is a more streamlined means of finding peak closing prices

#gafa_stock%>%
  #as.tibble()%>%
  #select(Symbol, Close)%>%
  #group_by(Symbol)%>%
  #summarise(maxi=max(Close))%>%
  #arrange(desc(maxi))%>%
  #flextable()
  


```

# Time Series Graphics: Q3

**Download the file tute1.csv from the book website, open it in Excel (or some other spreadsheet application), and review its contents.**

-   Read the data into R
-   Convert the data to time series
-   Construct time series plots of each of the three series
-   Check what happens when you don't include facet_grid().

*Without facet_grid the three series are plotted on the same graph with a continuous scale for the y-axis. This compresses the plotted data (i.e., reduced amplitude) and, in turn, reduces an investigators ability to distinguish fine-grained patterns within and across series.*

```{r}

# read data tute1 into R

tute1 <- read_csv("tute1.csv")

# convert tute1 into a time series

series <- tute1 %>%
  mutate(Quarter = yearmonth(Quarter)) %>% #tsibble represent year-month
  as_tsibble(index = Quarter) # date-times should be declared as index

# Construct time series plots of each of the three series

series %>%
  pivot_longer(-Quarter) %>%
  ggplot(aes(x = Quarter, y = value, colour = name)) +
  geom_line() +
  facet_grid(name ~ ., scales = "free_y")



```

# Time Series Graphics: Q4

**The USgas package contains data on the demand for natural gas in the US.**

-   Install the USgas package.
-   Create a tsibble from us_total with year as the index and state as the key.
-   Plot the annual natural gas consumption by state for the New England area (comprising the states of Maine, Vermont, New Hampshire, Massachusetts, Connecticut and Rhode Island).

*Note: USgas consists of three data sets:us_monthly, us_total, us_residential.*

```{r}

# create tsibble from us_total with year as the index and state as the key

total <- USgas::us_total%>%
  as_tsibble(key=state, index = year)

#Plot the annual natural gas consumption by state for the New England area (comprising the states of Maine, Vermont, New Hampshire, Massachusetts, Connecticut and Rhode Island).

total%>%
  filter(state %in% c('Maine', 'Vermont', 'New Hampshire', 'Massachusetts', 'Connecticut', 'Rhode Island'))%>%
  mutate_at(vars(y),funs(y=y/1000))%>% # data = units of million cubic feet: divide by 1000 for rendering
  autoplot()+
  labs(title= 'Annual Natural Gas Consumption in New England', y = 'Million cubic feet/1000', x= 'Year')+
  theme_classic()+
  facet_grid(state ~ ., scales = "free_y")

```

# Time Series Graphics: Q5

**Download tourism.xlsx from the book website and read it into R using readxl::read_excel().**

-   Create a tsibble which is identical to the tourism tsibble from the tsibble package.
-   Find what combination of Region and Purpose had the maximum number of overnight trips on average.
-   Create a new tsibble which combines the Purposes and Regions, and just has total trips by State.

*The combination of Region State with the maximum average number of overnight trips was Melbourne Visiting = 985.*

*Other combinations are included in Table 5.1 below.*

```{r}

# make tsibble identical to tsibble::tourism
# we need to change Quarter col to Date when importing Excel

tourism<-read_excel('tourism.xlsx')
tourism$Quarter<-as.Date(tourism$Quarter)

tourism%>%
  as_tsibble(key=c(Region, State,  Purpose), index = Quarter)%>%
  mutate(Quarter = yearquarter(Quarter)) 

# Combination of Region and Purpose with maximum number of overnight trips on average
  
aggregate(Trips ~ Region+Purpose, tourism, max)%>%
  group_by(Region)%>%
  arrange(desc(Trips))%>%
  slice(1)%>%
  arrange(desc(Trips))%>%
  flextable()%>%
  set_caption("Table 5.1: Maximum average number of overnight trips")
  
# total trips by state 

trips_state<-aggregate(Trips~Quarter+State, tourism, sum)%>%
    arrange(State)%>%
    as_tsibble(key=State, index = Quarter)

#plot total trips for visual assessment

trips_state%>%
  autoplot(.vars=Trips)+
  labs(title= 'Total Trips by State: 2000-2015', y = 'Number of Trips', x= 'Year')+
  theme_classic()+
  facet_grid(State ~ ., scales = "free_y")

```

# Time Series Graphics: Q8

**Monthly Australian retail data is provided in aus_retail. Select one of the time series as follows (but choose your own seed value):**

set.seed(12345678) myseries \<- aus_retail %>% filter(`Series ID` == sample(aus_retail\$`Series ID`,1))

**Explore your chosen retail time series using the following functions:**

-   autoplot()
-   gg_season()
-   gg_subseries()
-   gg_lag()
-   ACF()

**Can you spot any seasonality, cyclicity and trend? What do you learn about the series?**

*See response below following the graphs.*

```{r}
set.seed(124566791)

# load data

series <- tsibbledata::aus_retail %>%
  filter(`Series ID` == sample(aus_retail$`Series ID`,1)) #1 specifies number of Series ID groups

# autoplot

p1<-series%>%
  autoplot(Turnover)+
  labs(title = 'Figure 1. Inter-Annual Turnover in Pharmaceutical, Cosmetic and Toiletry Retail Industry', 
       subtitle = '1982-2020')+
  theme(axis.text.x = element_text(angle = 90))+
  theme_classic()

p2<-series%>%
  separate(Month, into=c('Year', 'Month'), sep=' ')%>%
  mutate(Year = as.numeric(Year))%>%
  filter(Year >= 1982)%>%
  ggplot(aes(Month, Turnover))+
  geom_point(aes( color=Year))+
  theme(axis.text.x = element_text(angle = 90))+
  labs(title = 'Figure 2. Monthly Turnover in Pharmaceutical, Cosmetic and Toiletry Retail Industry',
       subtitle = '1982-2020')+
  theme_classic()

# gg_season

p3<-series%>%
  gg_season(Turnover)+
  labs(title = 'Figure 3. Turnover in Pharmaceutical, Cosmetic and Toiletry Retail Industry',
       subtitle='By Month and Year')+
  theme_classic()

# gg_subseries

p4<-series%>%
  gg_subseries(Turnover)+
  labs(title = 'Figure 4. Turnover in Pharmaceutical, Cosmetic and Toiletry Retail Industry',
       subtitle='Subplots: Month and Year', x='Year')+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90))

# ACF

p5<-series%>%
  ACF(Turnover, lag_max=48) %>%
  autoplot()+
  labs('Figure 5. Autocorrelation: Turnover in Pharmaceutical, Cosmetic and Toiletry Retail Industry')+
  theme_classic()

# gg_lag

p6<-series%>%
  gg_lag(Turnover, geom = 'point', lags = 1:12)+
  labs(title = 'Figure 6. Lag in Turnover in Pharmaceutical, Cosmetic and Toiletry Retail Industry',
       x = "lag(Turnover, k)")+
  theme_classic()

# print plots

p1
p2
p3
p4
p5
p6



 
```

**Q8 Summary:**

The following patterns are apparent in this series:

-   *There is a strong increasing, non-stationary, trend (curvilinear) in Turnover between 1990-1920, which may include an ~ decadal pattern (starting in 1990) of increase followed by short decline. Fig. 1 & 4. A similar, less pronounced, cyclic pattern appears at the mid-point of the decadal period (Fig. 4).*
-   *There is a strong component of seasonality with turnover counts greatest in Dec and lowest in early winter (Jan-Feb). Fig. 2-4.*
-   *The amplitude of seasonality increases within and across years after ~ 2008. This may be a response to the 2008 recession (Fig. 1 & 4).*
-   *There is significant autocorrelation at every lag (Fig. 5 & 6) as well as trend and signs of seasonality (scalloping), with peaks 12 months apart (Fig. 6).*

*Retail turnover ($Million AUD) in the Pharmaceutical, Cosmetic, and Toiletry industry has increased since the early 1980's, with greater seasonal and inter-annual variation subsequent to \~2008. There is a strong seasonal component with turnover spiking with the end-of-year holiday season, followed by a lull for several months. The possibility of a decadal cycle in the industry (if confirmed) is interesting and warrants further investigation.*
